NURM: a tool to locate numerical differences in pipelines

### 2019-11-06 paper.tex
update result images and add discussion

Furthermore, different frequencies show the variability between subjects.
Results show that the same process can be reproducible for some subjects but not for others. 
Between subject differences can be a result of different subject types. 
The similar findings in the other studies show that the subject 
is an important factor in reproducibility~\cite{bowring2019exploring}.

% make figure in two columns use {figure*}
\begin{figure}
\centering
  \includegraphics[width=\columnwidth,height=4cm]{images/pfs_heatmap.png}
  \caption{Heatmap frequency of processes along with the sub-pipelines that introduce 
           differences among 20 subjects. 
           Each column represents the pipeline tool, and each row shows the sub-pipeline.
           Each cell may contain several process instances of a tool in a sub-pipeline.
           Annotations describe the frequency distribution of the instances as ``N [M] x Y" 
           in which N is the percentage of the process 
           that creates differences out of all M times that is occurred in the sub-pipeline. 
           Also, times Y shows the number of instances with the same distribution.
           For example, the cell of FLIRT tool in sub-pipeline AAve shows 10 process instances  
           with the frequency of 10 that introduce no differences.
           The range between red and green colors shows the highest and lowest 
           proportion of occurrences of the processes with differences.
           Also, grey cells indicate that the pipeline tool is not executed in the sub-pipeline.}
  \label{fig:pfs_freq}
\end{figure}

@@

\begin{figure}
  \centering
    \includegraphics[width=\columnwidth]{images/t2w_alignment.png} 
    \caption{Differences between FNIRT results of one subject from PreFreeSurfer 
    (checkerboard image \emph{T2w\_acpc\_to\_MNI\_nonli.nii.gz}) (CentOS6 vs. 
    CentOS7). Sagittal (top) and axial (bottom) views of the brain in dimension of X=130, Y=155, Z=130.
    Right images show the zoomed version part of the images marked by the red rectangle.
    Here is 
    \href{https://github.com/ali4006/HCP-reproducibility-paper/blob/master/images/pfs_t2w_alignment.gif}
    {the link} to the gif animation of the same images.
} 
    \label{fig:fnirt_result}
\end{figure}

@@

Figure~\ref{fig:tissue_class} illustrates the sum of binarized differences 
from 20 subject results of the brain tissue segmentation across OSes.
This figure shows between-OS differences  
as a result of the propagation and amplification of differences in FreeSurfer. 
Differences are mostly localized at the interfaces between tissues.

Furthermore, Figure~\ref{fig:scatter_plot} plots the mean of Dice coefficients 
for the 45 segmented regions obtained on two operating systems.
High Dice coefficients (more than 0.95) in most of the regions indicate minor differences.
Dice values can be very low , down to 0.5, in particular in small structures 
such as the vessels and lateral ventricles.

\begin{figure}
%  \includegraphics{brain\_classification}
\centering
  \includegraphics[width=\columnwidth]{images/brain_segmentation_mni.png} 
  \caption{Heatmap binarized differences between brain segmentation results (file \emph{aseg.hires.mgz}) from 
          20 subjects in FreeSurfer (CentOS6 vs. CentOS7). 
          All binarized differences were resampled to the default MNI152 volume template.
          Here is 
          \href{https://github.com/ali4006/HCP-reproducibility-paper/blob/master/images/fs_brain_segmentation.gif}
          {the link} to the gif animation from a single subject.} 
  \label{fig:tissue_class}
\end{figure}

\begin{figure}
  \centering
    \includegraphics[width=\columnwidth]{images/scatter_plot.png} 
    \caption{The scatter plot of the mean of Dice coefficients between segmented regions 
            from 20 subjetcs in FreeSurfer (CentOS6 vs. CentOS7). 
            The complete Dice coefficient values are 
            \href{https://github.com/ali4006/HCP-reproducibility-paper/blob/master/data/dice_values.csv}
            {available}.} 
    \label{fig:scatter_plot}
  \end{figure}

@@

We identified four processes that create differences 
among all 28 distinct processes that are involved in the PreFreeSurfer execution. 
The numerical instability in the 
PreFreeSurfer HCP pipeline arises mainly from linear 
registration processes implemented in FSL \emph{FLIRT}. 
We can see that \emph{FLIRT} creates differences in almost all the sub-pipelines. 
Therefore, it is important to investigate the registration procedure and find 
a way to stabilize this process.

In the second experiment, 
the analysis results obtained by the pipeline show that between-OS differences are visually substantial.
We see that statically builds of the pipeline improves reproducibility, but 
it would not improve the numerical stability of the pipeline. 
FreeSurfer pipeline propagates and amplifies the numerical differences that are created by the 
differences in the underlying operating system libraries.

In addition to variations of the operating system, other factors exist which may influence 
the reproducibility of the analyses such as software versions, compilers, hardware, etc.
Generally, we can use the Monte Carlo Arithmetic (MCA) to determine the numerical stability of the pipelines.
This algorithm perturbs every floating-point operation 
in the analysis based on the random rounding in each run. 
By this algorithm, NURM-tool can easily locate which parts of the analyzed source code are responsible 
for the floating-point related instabilities.

Although we identify the processes in the pipeline that create differences, 
the amplification of these differences should be quantified in the future works.

\note{FREESURFER RESULTS: 
These are consistant results to the paper~\cite{Glatard2015}, as it is shown the propaggation of differences in cortical thickness extraction analysis.
The magnitude of differences from the operating system changes is out of our expectaion}

@@

The preferred solution is to find the origin of amplification of the differences and 
then fixing the particular functions/modules in the processing pipelines.

>>>>

Furthermore, different frequencies show the variability between subjects.
Results show that the same process can be reproducible for some subjects but not for others. 
Between subject differences can be a result of different subject types. 
For instance, the subjects contain two T1 images need more calculations 
to produce an average image between them.
The similar findings in another study~\cite{bowring2019exploring} were reported 
the effect of subject type on reproducibility.

% make figure in two columns use {figure*}
\begin{figure*}
\centering
  \includegraphics[width=\textwidth]{images/pfs_heatmap.png}
  \caption{Heatmap frequency of processes along with the sub-pipelines that introduce 
           differences among 20 subjects. 
           Each column represents the pipeline tool, and each row shows the sub-pipeline.
           Each cell may contain several process instances of a tool in a sub-pipeline.
           Annotations describe the frequency distribution of the instances as ``N [M] x Y" 
           in which N is the percentage of the process 
           that creates differences out of all M times that is occurred in the sub-pipeline. 
           Also, times Y shows the number of instances with the same distribution.
           For example, the cell of FLIRT tool in sub-pipeline AAve shows 10 process instances  
           with the frequency of 10 that introduce no differences.
           The range between red and green colors shows the highest and lowest 
           proportion of occurrences of the processes with differences.
           Also, grey cells indicate that the pipeline tool is not executed in the sub-pipeline.}
  \label{fig:pfs_freq}
\end{figure*}

@@

\begin{figure}
  \centering
    \includegraphics[width=\columnwidth]{images/t2w_alignment.png} 
    \caption{Differences between FNIRT results of one subject from PreFreeSurfer 
    (checkerboard image \emph{T2w\_acpc\_to\_MNI\_nonli.nii.gz}) (CentOS6 vs. 
    CentOS7). Sagittal (left) and coronal (right) views of the brain are in dimension of X=130, Y=155, Z=130.
    \href{https://github.com/ali4006/HCP-reproducibility-paper/blob/master/images/pfs_t2w_alignment.gif}
    {the link} to the animated version of alignment results shows the better visualization of differences.
} 
    \label{fig:fnirt_result}
\end{figure}

@@

Figure~\ref{fig:tissue_class} illustrates the sum of binarized differences 
from 20 subject results of the brain tissue segmentation across OSes.
This figure shows between-OS differences  
as a result of the propagation and amplification of differences in FreeSurfer. 
Differences are mostly localized at the interfaces between tissues.
In our dataset, we found subjects with the min and max Dice coefficient 
values of 0.89 and 0.96 respectively.

Furthermore, Figure~\ref{fig:scatter_plot} plots the Dice coefficients 
for the 44 segmented regions obtained on the two operating systems.
High Dice coefficients (more than 0.95) in most of the regions indicate minor differences.
Dice values can be very low, down to 0.5, in particular in small structures 
such as the vessels and lateral ventricles.

\begin{figure}
%  \includegraphics{brain\_classification}
\centering
  \includegraphics[width=\columnwidth]{images/brain_segmentation_mni.png} 
  \caption{Heatmap binarized differences between brain segmentation results (file \emph{aseg.hires.mgz}) from 
          20 subjects in FreeSurfer (CentOS6 vs. CentOS7). 
          All binarized differences were resampled to the default MNI152 volume template.
          \href{https://github.com/ali4006/HCP-reproducibility-paper/blob/master/images/fs_brain_segmentation.gif}
          {The link} to the animated version of segmentation results of one subject represents the significance of differences.} 
  \label{fig:tissue_class}
\end{figure}

\begin{figure*}
  \centering
    \includegraphics[width=\textwidth]{images/scatter_plot.png} 
    \caption{The scatter plots of the Dice coefficients between segmented regions 
            from 20 subjects in FreeSurfer (CentOS6 vs. CentOS7). 
            The x-axes are the average number of voxels in each region.
            Regions are divided into four sub-plots from the smallest regions in the top-left plot 
            to the largest ones in the bottom-right plot.
            The complete table of Dice coefficient values is 
            \href{https://github.com/ali4006/HCP-reproducibility-paper/blob/master/data/dice_values.csv}
            {available}.} 
    \label{fig:scatter_plot}
  \end{figure*}

@@

We identified four processes that create differences 
among all 28 distinct processes that are involved in the PreFreeSurfer execution. 
The numerical instability in the 
PreFreeSurfer HCP pipeline arises mainly from linear 
registration processes implemented in FSL \emph{FLIRT}. 
We can see that \emph{FLIRT} creates differences in almost all the sub-pipelines. 
We monitored the library system calls using \emph{ltrace} to check that registration processes
use the same libraries. There were no differences in the library dependencies used by 
the tool in two operating systems, but we found a different number of calls to the \emph{glibc}.
More investigations need to stabilize registration processes.

In the second experiment, 
the analysis results obtained by the pipeline show that between-OS differences are visually substantial.
The statically builds of the pipeline would not improve the numerical stability of the pipeline. 
FreeSurfer pipeline propagates and amplifies the numerical differences that are created by the 
differences in the underlying operating system libraries.
These results are consistent with the findings in the paper~\cite{Glatard2015}, 
as it has confirmed the propagation of differences in cortical thickness extraction analysis using FreeSurfer.

We obtained results between-OS versions; further investigations could be done over the OS families (e.g., Linux VS. Windows).
In addition to variations of the operating system, numerical instability of the pipelines have been 
explored when varying software, hardware and implementation details, or by adding negligible perturbations.
In the field of numerical analysis, we can investigate these issue using the Monte Carlo Arithmetic (MCA), 
a method instrumenting floating-point operations based on the random rounding algorithms at a target precision.
By this algorithm, we can locate the instabilities for the given pipeline, tool, or operations.

Although we identify the processes in the pipeline that create differences, 
the amplification of these differences should be quantified in the future works.
We can use distance-preserving hashes rather than MD5 checksums or metrics specific to the file type like Dice coefficient 
for the registration processes to measure the magnitude of differences.

@@

\note{The preferred solution is to find the origin of amplification of the differences and 
then fixing the particular functions/modules in the processing pipelines. 
Bagging is a fundamental technique toward reducing the instability of the analysis.
Therefore, we can use bagging for the parts in the pipeline that amplify differences, 
but it still is computationaly expensive. }
